---
title: "Track options reference"
author:
- name: Arkadiusz Gladki
package: igvShiny
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Track options reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
    )
```

```{r load-package, results="hide", message=FALSE, warning=FALSE}
library(igvShiny)
```

# What this vignette is

Every track type `igvShiny` can load, and the options each one accepts. If you
are looking for how to get a browser onto the page in the first place, start
with the *Getting started with igvShiny* vignette instead.

The options are igv.js options: `igvShiny` passes them through rather than
reinventing them, and the names below are the ones the bundled igv.js
(currently 3.8.4) actually reads.

# How an option reaches the browser

Every loader takes a `trackConfig` argument: a named list merged into the track
configuration sent to igv.js.

```{r trackconfig, eval=FALSE}
loadBedTrack(
    session,
    id = "igv",
    trackName = "my regions",
    tbl = tbl,
    trackConfig = list(displayMode = "EXPANDED", maxRows = 20)
    )
```

The frequently used options are also plain arguments on the loaders that need
them — `color`, `trackHeight`, `displayMode` and so on. Both routes end up in
the same place, so use whichever reads better. Giving the same option twice is
not an error, but the explicit argument wins and you are told so:

```
Warning: User-provided trackConfig options conflict with function arguments
and will be ignored: displayMode
```

Keys are checked against an allowlist before they are sent, and a key that is
not on it is dropped:

```
Warning: Ignoring invalid or unsupported track options in trackConfig: onClick
```

That is the mechanism, not a nuisance: a `trackConfig` is a JavaScript object
built from user input, and only names known to be data options are let through.
The consequence to remember is that a typo does not reach igv.js and does not
raise an error — it produces a warning and a track drawn with defaults.

# Loaders

| Function | Data comes from | igv.js track |
|---|---|---|
| `loadBedTrack()` | `data.frame` | annotation |
| `loadBedGraphTrack()` | `data.frame` | wig |
| `loadBedGraphTrackFromURL()` | URL | wig |
| `loadSegTrack()` | `data.frame` | seg |
| `loadGwasTrack()` | `data.frame` | gwas |
| `loadVcfTrack()` | `VCF` object | variant |
| `loadGFF3TrackFromLocalData()` | `data.frame` | annotation |
| `loadGFF3TrackFromURL()` | URL | annotation |
| `loadBamTrackFromLocalData()` | `GAlignments` | alignment |
| `loadBamTrackFromURL()` | URL | alignment |
| `loadCramTrackFromLocalData()` | file on disk | alignment |
| `loadCramTrackFromURL()` | URL | alignment |
| `loadSpliceJunctionTrackFromURL()` | URL | junction |

The `FromURL` variants ask the *user's browser* to fetch the file. A URL that
works in `download.file()` can still fail here if the server sends no
permissive CORS header — see the troubleshooting section of the getting started
vignette.

# Options every track accepts

| Option | Type | What it does |
|---|---|---|
| `name` | character | track label in the left panel |
| `url` | character | where the data is |
| `indexURL` | character | index for a bgzipped or binary file |
| `indexed` | logical | set `FALSE` to read an unindexed file whole |
| `format` | character | `"bed"`, `"gff3"`, `"bigwig"`, `"vcf"`, … |
| `type` | character | igv.js track class; usually implied by the loader |
| `order` | numeric | position among the other tracks |
| `height` | numeric | track height in pixels |
| `minHeight`, `maxHeight` | numeric | bounds when the track resizes itself |
| `autoHeight` | logical | grow to fit the features in view |
| `visibilityWindow` | numeric | above this span in bp, draw nothing |
| `removable` | logical | whether the user may close the track |
| `color` | character | any CSS colour |
| `altColor` | character | second colour, per track type |
| `displayMode` | character | `"COLLAPSED"`, `"EXPANDED"`, `"SQUISHED"` |
| `roi` | list | regions of interest drawn over this track |
| `oauthToken`, `headers` | character, list | sent with the data request |

`trackHeight` on the loaders and `height` in `trackConfig` are the same thing.

# Numeric tracks: bedGraph, wig, bigWig

| Option | Type | What it does |
|---|---|---|
| `autoscale` | logical | rescale to the data in view |
| `autoscaleGroup` | character | scale several tracks together, as one group |
| `min`, `max` | numeric | fixed data range when not autoscaling |
| `logScale` | logical | log the y axis |
| `graphType` | character | `"bar"` or `"points"` |
| `flipAxis` | logical | draw the y axis upside down |
| `color`, `altColor` | character | values above and below the baseline |

```{r wig, eval=FALSE}
loadBedGraphTrack(
    session, id = "igv", trackName = "coverage", tbl = tbl,
    trackConfig = list(
        graphType = "points",
        autoscaleGroup = "sampleA",
        logScale = FALSE
        )
    )
```

Two tracks sharing an `autoscaleGroup` are drawn on one scale, which is what
makes their heights comparable by eye.

# Annotation tracks: bed, gff3

| Option | Type | What it does |
|---|---|---|
| `colorBy` | character | feature attribute to take the colour from |
| `colorTable` | list | attribute value to colour |
| `featureHeight` | numeric | height of one feature row |
| `maxRows` | numeric | rows drawn before the rest are hidden |
| `searchable` | logical | let the locus box find features by name |
| `queryable` | logical | whether the track answers region queries |

On `loadGFF3TrackFromURL()` and `loadGFF3TrackFromLocalData()` the argument is
called `colorByAttribute`, and it becomes igv.js `colorBy`. Pass it as the
loader argument rather than in `trackConfig`, together with `colorTable`:

```{r gff3, eval=FALSE}
loadGFF3TrackFromURL(
    session, id = "igv", trackName = "genes",
    gff3URL = url, indexURL = paste0(url, ".tbi"),
    colorByAttribute = "biotype",
    colorTable = list(
        protein_coding = "darkgreen",
        processed_transcript = "blue",
        default = "black"
        )
    )
```

A `default` entry in `colorTable` catches the values you did not list.

# Alignment tracks: bam, cram

| Option | Type | What it does |
|---|---|---|
| `showAllBases` | logical | draw every base, not just mismatches |
| `viewAsPairs` | logical | join mates on one row |
| `colorBy` | character | `"strand"`, `"firstOfPairStrand"`, `"tag"`, … |
| `colorTable` | list | value to colour, for `colorBy = "tag"` |
| `sort` | list | how reads are sorted when the track loads |
| `samplingWindowSize` | numeric | window the downsampler works over |
| `samplingDepth` | numeric | reads kept per window |

`sort` takes the same object the igv.js right-click menu builds, so sorting by
a tag at load time is:

```{r bam-sort, eval=FALSE}
loadBamTrackFromURL(
    session, id = "igv", trackName = "reads",
    bamURL = bam, indexURL = paste0(bam, ".bai"),
    trackConfig = list(
        sort = list(
            option = "TAG", tag = "HP",
            chr = "chr8", position = 128750000
            )
        )
    )
```

Alignment tracks are the memory-hungry ones. `visibilityWindow` matters here
more than anywhere else: without it a user who zooms out asks the browser for
every read on the chromosome.

# Variant tracks: vcf

| Option | Type | What it does |
|---|---|---|
| `colorBy` | character | variant attribute to colour by |
| `colorTable` | list | attribute value to colour |
| `maxRows` | numeric | genotype rows drawn |
| `sort` | list | initial sort of the genotype rows |
| `displayMode` | character | `"COLLAPSED"` hides the genotypes |

# GWAS tracks

| Option | Type | What it does |
|---|---|---|
| `trait` | character | column holding the trait name |
| `columns` | list | **1-based** column positions in the source file |
| `min`, `max` | numeric | y range, in -log10(p) |
| `autoscale` | logical | rescale to the data in view |
| `colorTable` | list | chromosome to colour |

`columns` is worth its own note. The GWAS parser reads it 1-based, and without
it the parser guesses the layout from header names it recognises — any other
spelling draws an empty track rather than an error:

```{r gwas, eval=FALSE}
loadGwasTrack(
    session, id = "igv", trackName = "gwas", tbl.gwas = tbl,
    trackConfig = list(
        columns = list(chromosome = 12, position = 13, value = 28)
        )
    )
```

The `GWASTrack` class is the other way into the same track type, for data that
already lives at a URL; see `?GWASTrack`.

# Splice junction tracks

Junction options divide into filters, which decide whether an arc is drawn at
all, and appearance.

Filters:

| Option | Type | What it does |
|---|---|---|
| `minUniquelyMappedReads` | numeric | drop junctions below this count |
| `minTotalReads` | numeric | unique plus multi-mapped |
| `maxFractionMultiMappedReads` | numeric | drop mostly multi-mapped junctions |
| `minSplicedAlignmentOverhang` | numeric | shortest anchor accepted |
| `minJunctionEndsVisible` | numeric | 0, 1 or 2 ends in view |
| `minSamplesWithThisJunction` | numeric | across samples |
| `maxSamplesWithThisJunction` | numeric | across samples |
| `minPercentSamplesWithThisJunction` | numeric | as a percentage |
| `maxPercentSamplesWithThisJunction` | numeric | as a percentage |
| `hideAnnotatedJunctions` | logical | draw novel junctions only |
| `hideUnannotatedJunctions` | logical | draw annotated junctions only |
| `hideMotifs` | character | motifs to leave out, e.g. `c("GT/AG")` |
| `hideStrand` | character | `"+"` or `"-"` |

Appearance:

| Option | Type | What it does |
|---|---|---|
| `thicknessBasedOn` | character | `"numUniqueReads"`, `"numReads"`, `"isAnnotatedJunction"` |
| `bounceHeightBasedOn` | character | `"random"`, `"distance"`, `"thickness"` |
| `colorBy` | character | `"numUniqueReads"`, `"numReads"`, `"isAnnotatedJunction"`, `"strand"`, `"motif"` |
| `colorByNumReadsThreshold` | numeric | split point for the read-count colouring |
| `labelWith` | character | what the arc label shows |
| `labelWithInParen` | character | a second value, in parentheses |

```{r junctions, eval=FALSE}
loadSpliceJunctionTrackFromURL(
    session, id = "igv", trackName = "sampleA junctions",
    url = junctions.url, indexURL = paste0(junctions.url, ".tbi"),
    trackHeight = 150,
    trackConfig = list(
        colorBy = "motif",
        labelWith = "uniquelyMapped",
        minUniquelyMappedReads = 5,
        hideAnnotatedJunctions = FALSE
        )
    )
```

The filter names come from the igv.js source rather than from the
`spliceJunctionTrack.html` example page that ships with it: that example sets
`labelUniqueReadCount` and four siblings, and no such option exists in the
library any more.

# Options on startup tracks

Tracks passed to `igvShiny()` through `tracks` are checked against the same
allowlist, so everything above applies to them too. What they additionally need
is a `url`, since there is no loader argument to carry the data: an entry
without a usable one is dropped with a warning.

```{r startup, eval=FALSE}
output$igv <- renderIgvShiny({
    igvShiny(
        genomeOptions,
        tracks = list(
            list(
                name = "genes",
                type = "annotation",
                format = "gff3",
                url = gff3.url,
                indexed = FALSE,
                displayMode = "EXPANDED"
                )
            )
        )
    })
```

# When an option seems to do nothing

**Check for the warning first.** A dropped key says so on the R console. If
there is no warning, the option reached igv.js and the problem is elsewhere.

**An option igv.js does not know is ignored in silence** once past the
allowlist. igv.js reads the keys it recognises and never complains about the
rest, so a correct-looking option from IGV desktop, or from an older igv.js,
produces a track drawn with defaults and no message anywhere.

**Some options apply to one track type only.** `autoscale` on an annotation
track, `colorBy = "motif"` on anything that is not a junction track: both are
accepted, and both do nothing.

When you need an option this vignette does not list, the
[igv.js wiki](https://github.com/igvteam/igv.js/wiki) documents the underlying
browser. If it is genuinely missing from the allowlist rather than misspelled,
that is worth [an issue](https://github.com/gladkia/igvShiny/issues) — the list
covers the options that are useful from R, and it grows.

# Session Info

```{r session-info}
sessionInfo()
```
